Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile

### 2024-01-16 article/paper_drought_Chile2023.qmd
update article landscape figures of trend

  Human-induced greenhouse gas emissions have increased the frequency and/or intensity of weather and climate extremes. Central Chile has been affected by a persistent drought which is impacting the hydrological system and vegetation development. The region has been the focus of research studies due to the diminishing water supply, this persistent period of water scarcity has been defined as a “mega drought". Nevertheless, our results evidence that the water deficit has expanded beyond. Our goal is to analyze the impact of drought, measured by drought indices of water supply/demand and vegetation status, in the LULCC (land use land cover change) over continental Chile.  For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande”,  “Norte Chico”,  “Zona Central”, “Zona Sur”, and “Zona Austral”.  We used monthly climatic re-analysis variables for precipitation, temperature and soil moisture for 1981-2023 from ERA5-Land; and MODIS (Moderate-Resolution Imaging Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023, both from collection 6.1. We estimated atmospheric evaporative demand (AED) by combining the Hargreaves-Samani equation with the ERA5-Land temperature. We derived the drought indices SPI (Standardized Precipitation Index), SPEI (Standardized Precipitation Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), zcSM (standardized anomaly of cumulative soil moisture), and the zcNDVI (standardized anomaly of cumulative NDVI). These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1, 3, and 6 months). We analyzed the temporal correlation of SPI, SPEI, EDDI, and zcSM with zcNDVI to have insights into the impact of water supply and demand on vegetation.   Our results showed that LULCC had an increasing trend of 412 [km2yr−1] of forest expansion in the “Zona Sur”, together with a  decreasing trend of 24 [km2yr−1] of cropland contraction in the “Zona Central” meanwhile the “Zona Sur” showed an increase of 31 [km2yr−1], and a contraction of 80 [km2yr−1] of bare soil in the “Zona Austral”. The EDDI was the less correlated index for the five macro zones and the five types of land cover, showing that the temperature in Chile has little impact on vegetation. Higher r-squared values, between 0.5 and 0.8, were obtained at “Norte Chico” and “Zona Central” for the land cover types of savanna, shrubland, grassland, and croplands for the indices SPEI and zcSM at time scales of 12 and 24 months. The forest type reaches a r-squared of ~0.5 for zcSM of 12 months. The results indicate that the “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits longer than a year, potentially explained by a low capacity of water storage in those zones that should be further investigated.

@@

      text: '\usepackage{lineno}\linenumbers
             \usepackage{multirow}'

@@

```{r fig-studyArea}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: '(a) Chile with the Koppen-Geiger climate classes and the five macrozones "Norte Grande", "Norte Chico",  "Centro", "Sur", and "Austral". (b) Topography reference map. (c) Land cover classes for 2022. (d) Persistent land cover classes (> 80%) for 2001-2022'
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/map_study_con_landcover.png")
```

@@

For water supply and demand variabes, we used ERA5-Land [@MunozSabater2021], a reanalys dataset wich provides the evolution of land variables since 1950. It has a spatial resolution of 0.1°, hourly frequency and global coverage. We selected the variables for total precipitation, 2 meters temperature maximum and minimum, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).

To derive a proxy of vegetation productivity, we used the product MOD13A3 collection 6.1 from MODIS [@Didan2015]. It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency.

@@

To evaluate water demand, we chose the $EDDI$ [@Hobbins2016; @McEvoy2016] index, which uses the $AED$. For supply, we used the index recommended by the World Meteorological Organization (WMO) for monitoring drought, the SPI [@Mckee1993]. We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used. In this study, we propose the $zcSM$ (standardized anomaly of acumulated soil moisture at 1 m), which uses SM. Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI. Before using the NDVI, it was smoothed following the procedure described in @Zambrano2018 and @Zambrano2016.

@@

The drought indices $SPI$, $SPEI$, $EDDI$, $zcSM$, and $zcNDVI$ are obtained following the inverse normal approximation:

@@

To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975]. To determine the magnitude of the trend, we used Sen's slope [@Sen1968]. One of the advantages of applying this methodology is that the Sen's slope is not affected by outliers as regular regression does. We applied both to the five indices at the six time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and zcSM, and from 2000 to 2023 in the case of zcNDVI.

@@

To assess how water demand and supply and soil moisture are related to vegetation productivity (zcNDVI), we analyze the linear correlation between the indices SPI, SPEI, EDDI, and zcSM for 1, 3, 6, 12, 24, and 36-month time scales against zcNDVI. We followed a similar approach to that used by @Meroni2016 when using the SPI for meteorological drought against the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation) as a proxy for vegetation productivity. We made a pixel-to-pixel linear correlation analysis per index. First, we calculate the Pearson coefficient of correlation for the six time scales and let the time scale that reaches the maximum correlation be significant (p < 0.05). Then, we extracted the Pearson correlation value corresponding to the time scales that reached the maximum value. Thus, we derived two raster maps per index, the first with the time scales and the second with the correlation value.

@@

```{r fig-trendSPI}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "Linear trend of the SPI at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
#| out-width: "\\textwidth"

knitr::include_graphics("../output/figs/trend_raster_SPI_1981-2023.png")
```

```{r fig-trendEDDI}
#| echo: false
#| fig-pos: "!ht"
#| out-width: "\\textwidth"
#| fig-cap: "Linear trend of the EDDI at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
knitr::include_graphics("../output/figs/trend_raster_EDDI_1981-2023.png")
```

>>>>

  Human-induced greenhouse gas emissions have increased the frequency and/or intensity of weather and climate extremes. Central Chile has been affected by a persistent drought which is impacting the hydrological system and vegetation development. The region has been the focus of research studies due to the diminishing water supply, this persistent period of water scarcity has been defined as a “mega drought". Nevertheless, our results evidence that the water deficit has expanded beyond. Our goal is to analyze the impact of drought, measured by drought indices of water supply/demand and vegetation status, in the LULCC (land use land cover change) over continental Chile.  For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande”,  “Norte Chico”,  “Zona Central”, “Zona Sur”, and “Zona Austral”.  We used monthly climatic re-analysis variables for precipitation, temperature and soil moisture for 1981-2023 from ERA5-Land; and MODIS (Moderate-Resolution Imaging Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023, both from collection 6.1. We estimated atmospheric evaporative demand (AED) by combining the Hargreaves-Samani equation with the ERA5-Land temperature. We derived the drought indices SPI (Standardized Precipitation Index), SPEI (Standardized Precipitation Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI (standardized anomaly of cumulative soil moisture), and the zcNDVI (standardized anomaly of cumulative NDVI). These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1, 3, and 6 months). We analyzed the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to have insights into the impact of water supply and demand on vegetation.   Our results showed that LULCC had an increasing trend of 412 [km2yr−1] of forest expansion in the “Zona Sur”, together with a  decreasing trend of 24 [km2yr−1] of cropland contraction in the “Zona Central” meanwhile the “Zona Sur” showed an increase of 31 [km2yr−1], and a contraction of 80 [km2yr−1] of bare soil in the “Zona Austral”. The EDDI was the less correlated index for the five macro zones and the five types of land cover, showing that the temperature in Chile has little impact on vegetation. Higher r-squared values, between 0.5 and 0.8, were obtained at “Norte Chico” and “Zona Central” for the land cover types of savanna, shrubland, grassland, and croplands for the indices SPEI and SSI at time scales of 12 and 24 months. The forest type reaches a r-squared of ~0.5 for SSI of 12 months. The results indicate that the “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits longer than a year, potentially explained by a low capacity of water storage in those zones that should be further investigated.

@@

      text: \usepackage{lineno}\linenumbers
            \usepackage{multirow}
            \usepackage{lscape}
            \newcommand{\blandscape}{\begin{landscape}}
            \newcommand{\elandscape}{\end{landscape}}

@@

```{r fig-studyArea}
#| echo: false
#| fig-pos: "!ht"
#| fig-env: 'figure*'
#| fig-cap: '(a) Chile with the Koppen-Geiger climate classes and the five macrozones "Norte Grande", "Norte Chico",  "Centro", "Sur", and "Austral". (b) Topography reference map. (c) Land cover classes for 2022. (d) Persistent land cover classes (> 80%) for 2001-2022'
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/map_study_con_landcover.png")
```

@@

For water supply and demand variables, we used ERA5-Land [@MunozSabater2021], a reanalys dataset that provides the evolution of land variables since 1950. It has a spatial resolution of 0.1°, hourly frequency, and global coverage. We selected the variables for total precipitation, 2 meter temperature maximum and minimum, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3). The data was downloaded using the Copernicus Climate Data Store (CDS) Application Program Interface (API) implemented in `{ecmfwr}` [@Hufkens2019].

To derive a proxy of vegetation productivity, we used the product MOD13A3 collection 6.1 from MODIS [@Didan2015]. It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency. The MOD13A3.061 and MCD12Q1.061 were retrieved from the online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, https://lpdaa.usgs.gov/tools/data-pool/.

@@

To evaluate water demand, we chose the $EDDI$ [@Hobbins2016; @McEvoy2016] index, which uses the $AED$. For supply, we used the index recommended by the World Meteorological Organization (WMO) for monitoring drought, the SPI [@Mckee1993]. We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used. In this study, we used the $SSI$ (standardized soil moisture index at 1 m) [@Hao2013;@AghaKouchak2014], which uses SM. Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI. Before using the NDVI, it was smoothed following the procedure described in @Zambrano2018 and @Zambrano2016.

@@

The drought indices $SPI$, $SPEI$, $EDDI$, $SSI$, and $zcNDVI$ are obtained following the inverse normal approximation:

@@

To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975]. To determine the magnitude of the trend, we used Sen's slope [@Sen1968]. One of the advantages of applying this methodology is that the Sen's slope is not affected by outliers as regular regression does. We applied both to the five indices at the six time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and SSI, and from 2000 to 2023 in the case of zcNDVI.

@@

To assess how water demand and supply and soil moisture are related to vegetation productivity (zcNDVI), we analyze the linear correlation between the indices SPI, SPEI, EDDI, and SSI for 1, 3, 6, 12, 24, and 36-month time scales against zcNDVI. We followed a similar approach to that used by @Meroni2016 when using the SPI for meteorological drought against the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation) as a proxy for vegetation productivity. We made a pixel-to-pixel linear correlation analysis per index. First, we calculate the Pearson coefficient of correlation for the six time scales and let the time scale that reaches the maximum correlation be significant (p < 0.05). Then, we extracted the Pearson correlation value corresponding to the time scales that reached the maximum value. Thus, we derived two raster maps per index, the first with the time scales and the second with the correlation value.

@@

\newpage

\blandscape
```{r fig-trendSPI}
#| echo: false
#| layout-ncol: 2
#| fig-pos: "!ht"
#| fig-env: "figure*"
#| fig-cap: "Linear trend of the drought index (*) at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
#| fig-subcap: 
#| - "SPI (Standardized Precipitation Index)"
#| - "SPEI (Standardized Precipitation Evapotranspiration Index)" 
#| - "EDDI (Evaporative Demand Drought Index)"
#| - "SSMI (Standardized Soil Moisture Index)"
#| out-width: "\\textwidth"

knitr::include_graphics("../output/figs/trend_raster_SPI_1981-2023.png")
knitr::include_graphics("../output/figs/trend_raster_SPEI_1981-2023.png")
knitr::include_graphics("../output/figs/trend_raster_EDDI_1981-2023.png")
knitr::include_graphics("../output/figs/trend_raster_zcSM_1981-2023.png")

```

\elandscape

### 2024-01-16 article/paper_drought_Chile2023.tex
update article landscape figures of trend

\usepackage{lineno}\linenumbers \usepackage{multirow}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Human-induced greenhouse gas emissions have increased the frequency
and/or intensity of weather and climate extremes. Central Chile has been
affected by a persistent drought which is impacting the hydrological
system and vegetation development. The region has been the focus of
research studies due to the diminishing water supply, this persistent
period of water scarcity has been defined as a ``mega drought''.
Nevertheless, our results evidence that the water deficit has expanded
beyond. Our goal is to analyze the impact of drought, measured by
drought indices of water supply/demand and vegetation status, in the
LULCC (land use land cover change) over continental Chile. For the
analysis, continental Chile was divided into five zones according to a
latitudinal gradient: ``Norte Grande'', ``Norte Chico'', ``Zona
Central'', ``Zona Sur'', and ``Zona Austral''. We used monthly climatic
re-analysis variables for precipitation, temperature and soil moisture
for 1981-2023 from ERA5-Land; and MODIS (Moderate-Resolution Imaging
Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the
NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023,
both from collection 6.1. We estimated atmospheric evaporative demand
(AED) by combining the Hargreaves-Samani equation with the ERA5-Land
temperature. We derived the drought indices SPI (Standardized
Precipitation Index), SPEI (Standardized Precipitation
Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), zcSM
(standardized anomaly of cumulative soil moisture), and the zcNDVI
(standardized anomaly of cumulative NDVI). These indices were calculated
for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1,
3, and 6 months). We analyzed the temporal correlation of SPI, SPEI,
EDDI, and zcSM with zcNDVI to have insights into the impact of water
supply and demand on vegetation. Our results showed that LULCC had an
increasing trend of 412 {[}km2yr−1{]} of forest expansion in the ``Zona
Sur'', together with a decreasing trend of 24 {[}km2yr−1{]} of cropland
contraction in the ``Zona Central'' meanwhile the ``Zona Sur'' showed an
increase of 31 {[}km2yr−1{]}, and a contraction of 80 {[}km2yr−1{]} of
bare soil in the ``Zona Austral''. The EDDI was the less correlated
index for the five macro zones and the five types of land cover, showing
that the temperature in Chile has little impact on vegetation. Higher
r-squared values, between 0.5 and 0.8, were obtained at ``Norte Chico''
and ``Zona Central'' for the land cover types of savanna, shrubland,
grassland, and croplands for the indices SPEI and zcSM at time scales of
12 and 24 months. The forest type reaches a r-squared of
\textasciitilde0.5 for zcSM of 12 months. The results indicate that the
``Norte Chico'' and ``Zona Central'' are the most sensitive regions to
water supply deficits longer than a year, potentially explained by a low
capacity of water storage in those zones that should be further
investigated.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden]}{\end{tcolorbox}}\fi

@@

@@

@@

For water supply and demand variabes, we used ERA5-Land
\citep{MunozSabater2021}, a reanalys dataset wich provides the evolution
of land variables since 1950. It has a spatial resolution of 0.1°,
hourly frequency and global coverage. We selected the variables for
total precipitation, 2 meters temperature maximum and minimum, and
volumetric soil water layers between 0 and 100cm of depth (layer 1 to
layer 3).

To derive a proxy of vegetation productivity, we used the product
MOD13A3 collection 6.1 from MODIS \citep{Didan2015}. It provides
vegetation indices (NDVI and EVI) at 1km of spatial resolution and
monthly frequency.

@@

To evaluate water demand, we chose the \(EDDI\)
\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). For
supply, we used the index recommended by the World Meteorological
Organization (WMO) for monitoring drought, the SPI \citep{Mckee1993}. We
calculated the SPEI, which used a balance between \(P\) and \(AED\), in
this case, an auxiliary variable \(D = P-AED\) is used. In this study,
we propose the \(zcSM\) (standardized anomaly of acumulated soil
moisture at 1 m), which uses SM. Finally, for the proxy of productivity,
\(zcNDVI\), we used the NDVI. Before using the NDVI, it was smoothed
following the procedure described in \citep{Zambrano2018} and
\citep{Zambrano2016}.

@@

The drought indices \(SPI\), \(SPEI\), \(EDDI\), \(zcSM\), and
\(zcNDVI\) are obtained following the inverse normal approximation:

@@

To estimate if there are significant positive or negative trends for the
drought indices, we used the non-parametric test of Mann-Kendall
\citep{Kendall1975}. To determine the magnitude of the trend, we used
Sen's slope \citep{Sen1968}. One of the advantages of applying this
methodology is that the Sen's slope is not affected by outliers as
regular regression does. We applied both to the five indices at the six
time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and zcSM, and
from 2000 to 2023 in the case of zcNDVI.

@@

To assess how water demand and supply and soil moisture are related to
vegetation productivity (zcNDVI), we analyze the linear correlation
between the indices SPI, SPEI, EDDI, and zcSM for 1, 3, 6, 12, 24, and
36-month time scales against zcNDVI. We followed a similar approach to
that used by \citep{Meroni2016} when using the SPI for meteorological
drought against the cumulative FAPAR (Fraction of Absorbed
Photosynthetically Active Radiation) as a proxy for vegetation
productivity. We made a pixel-to-pixel linear correlation analysis per
index. First, we calculate the Pearson coefficient of correlation for
the six time scales and let the time scale that reaches the maximum
correlation be significant (p \textless{} 0.05). Then, we extracted the
Pearson correlation value corresponding to the time scales that reached
the maximum value. Thus, we derived two raster maps per index, the first
with the time scales and the second with the correlation value.

@@

Analyzing the water supply in the macrozones ``norte chico,''
``central,'' and ``sur,'' the SPI shows a decreasing trend that
increases at longer time scales due to the prolonged reduction in
precipitation. It reaches a trend of -0.056 (z-score) per decade
(Figure~\ref{fig-trendSPI}) for time scales of 36 months. In the case of
water demand, we analyze the EDDI. It shows a positive trend, caused by
an increase in AED. The trend on EDDI reaches a maximum of 0.053 per
decade in the macrozones ``norte grande'' and ``norte chico''
(Figure~\ref{fig-trendEDDI}). The behavior is similar to SPI, having
close trends with opposite signs. Nevertheless, the spatial patterns are
different. The maximum trend for SPI-36 is in ``norte chico'' and ``zona
central.'' For EDDI-36, the maximum trend is from ``northe grande'' to
``zona central.''

{\centering \includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}

@@

\caption{\label{fig-trendSPI}Linear trend of the SPI at time scales of
1, 3, 6, 12, 24, and 36 months for 1981-2023}

@@

{\centering \includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}

\caption{\label{fig-trendEDDI}Linear trend of the EDDI at time scales of
1, 3, 6, 12, 24, and 36 months for 1981-2023}

>>>>

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Human-induced greenhouse gas emissions have increased the frequency
and/or intensity of weather and climate extremes. Central Chile has been
affected by a persistent drought which is impacting the hydrological
system and vegetation development. The region has been the focus of
research studies due to the diminishing water supply, this persistent
period of water scarcity has been defined as a ``mega drought''.
Nevertheless, our results evidence that the water deficit has expanded
beyond. Our goal is to analyze the impact of drought, measured by
drought indices of water supply/demand and vegetation status, in the
LULCC (land use land cover change) over continental Chile. For the
analysis, continental Chile was divided into five zones according to a
latitudinal gradient: ``Norte Grande'', ``Norte Chico'', ``Zona
Central'', ``Zona Sur'', and ``Zona Austral''. We used monthly climatic
re-analysis variables for precipitation, temperature and soil moisture
for 1981-2023 from ERA5-Land; and MODIS (Moderate-Resolution Imaging
Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the
NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023,
both from collection 6.1. We estimated atmospheric evaporative demand
(AED) by combining the Hargreaves-Samani equation with the ERA5-Land
temperature. We derived the drought indices SPI (Standardized
Precipitation Index), SPEI (Standardized Precipitation
Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI
(standardized anomaly of cumulative soil moisture), and the zcNDVI
(standardized anomaly of cumulative NDVI). These indices were calculated
for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1,
3, and 6 months). We analyzed the temporal correlation of SPI, SPEI,
EDDI, and SSI with zcNDVI to have insights into the impact of water
supply and demand on vegetation. Our results showed that LULCC had an
increasing trend of 412 {[}km2yr−1{]} of forest expansion in the ``Zona
Sur'', together with a decreasing trend of 24 {[}km2yr−1{]} of cropland
contraction in the ``Zona Central'' meanwhile the ``Zona Sur'' showed an
increase of 31 {[}km2yr−1{]}, and a contraction of 80 {[}km2yr−1{]} of
bare soil in the ``Zona Austral''. The EDDI was the less correlated
index for the five macro zones and the five types of land cover, showing
that the temperature in Chile has little impact on vegetation. Higher
r-squared values, between 0.5 and 0.8, were obtained at ``Norte Chico''
and ``Zona Central'' for the land cover types of savanna, shrubland,
grassland, and croplands for the indices SPEI and SSI at time scales of
12 and 24 months. The forest type reaches a r-squared of
\textasciitilde0.5 for SSI of 12 months. The results indicate that the
``Norte Chico'' and ``Zona Central'' are the most sensitive regions to
water supply deficits longer than a year, potentially explained by a low
capacity of water storage in those zones that should be further
investigated.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, boxrule=0pt, enhanced, sharp corners, interior hidden]}{\end{tcolorbox}}\fi

@@

\begin{figure*}[!ht]

@@

\end{figure*}

@@

For water supply and demand variables, we used ERA5-Land
\citep{MunozSabater2021}, a reanalys dataset that provides the evolution
of land variables since 1950. It has a spatial resolution of 0.1°,
hourly frequency, and global coverage. We selected the variables for
total precipitation, 2 meter temperature maximum and minimum, and
volumetric soil water layers between 0 and 100cm of depth (layer 1 to
layer 3). The data was downloaded using the Copernicus Climate Data
Store (CDS) Application Program Interface (API) implemented in
\texttt{\{ecmfwr\}} \citep{Hufkens2019}.

To derive a proxy of vegetation productivity, we used the product
MOD13A3 collection 6.1 from MODIS \citep{Didan2015}. It provides
vegetation indices (NDVI and EVI) at 1km of spatial resolution and
monthly frequency. The MOD13A3.061 and MCD12Q1.061 were retrieved from
the online Data Pool, courtesy of the NASA EOSDIS Land Processes
Distributed Active Archive Center (LP DAAC), USGS Earth Resources
Observation and Science (EROS) Center, Sioux Falls, South Dakota,
https://lpdaa.usgs.gov/tools/data-pool/.

@@

To evaluate water demand, we chose the \(EDDI\)
\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). For
supply, we used the index recommended by the World Meteorological
Organization (WMO) for monitoring drought, the SPI \citep{Mckee1993}. We
calculated the SPEI, which used a balance between \(P\) and \(AED\), in
this case, an auxiliary variable \(D = P-AED\) is used. In this study,
we used the \(SSI\) (standardized soil moisture index at 1 m)
\citep{Hao2013, AghaKouchak2014}, which uses SM. Finally, for the proxy
of productivity, \(zcNDVI\), we used the NDVI. Before using the NDVI, it
was smoothed following the procedure described in \citep{Zambrano2018}
and \citep{Zambrano2016}.

@@

The drought indices \(SPI\), \(SPEI\), \(EDDI\), \(SSI\), and \(zcNDVI\)
are obtained following the inverse normal approximation:

@@

To estimate if there are significant positive or negative trends for the
drought indices, we used the non-parametric test of Mann-Kendall
\citep{Kendall1975}. To determine the magnitude of the trend, we used
Sen's slope \citep{Sen1968}. One of the advantages of applying this
methodology is that the Sen's slope is not affected by outliers as
regular regression does. We applied both to the five indices at the six
time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and SSI, and
from 2000 to 2023 in the case of zcNDVI.

@@

To assess how water demand and supply and soil moisture are related to
vegetation productivity (zcNDVI), we analyze the linear correlation
between the indices SPI, SPEI, EDDI, and SSI for 1, 3, 6, 12, 24, and
36-month time scales against zcNDVI. We followed a similar approach to
that used by \citep{Meroni2016} when using the SPI for meteorological
drought against the cumulative FAPAR (Fraction of Absorbed
Photosynthetically Active Radiation) as a proxy for vegetation
productivity. We made a pixel-to-pixel linear correlation analysis per
index. First, we calculate the Pearson coefficient of correlation for
the six time scales and let the time scale that reaches the maximum
correlation be significant (p \textless{} 0.05). Then, we extracted the
Pearson correlation value corresponding to the time scales that reached
the maximum value. Thus, we derived two raster maps per index, the first
with the time scales and the second with the correlation value.

@@

Analyzing the water supply in the macrozones ``norte chico,''
``central,'' and ``sur,'' the SPI shows a decreasing trend that
increases at longer time scales due to the prolonged reduction in
precipitation. It reaches a trend of -0.056 (z-score) per decade
(Figure~\ref{fig-trendSPI}) for time scales of 36 months. In the case of
water demand, we analyze the EDDI. It shows a positive trend, caused by
an increase in AED. The trend on EDDI reaches a maximum of 0.053 per
decade in the macrozones ``norte grande'' and ``norte chico''
(\textbf{?@fig-trendEDDI}). The behavior is similar to SPI, having close
trends with opposite signs. Nevertheless, the spatial patterns are
different. The maximum trend for SPI-36 is in ``norte chico'' and ``zona
central.'' For EDDI-36, the maximum trend is from ``northe grande'' to
``zona central.''

\newpage

\blandscape

\begin{figure}

\begin{minipage}[t]{0.50\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}

@@

\subcaption{\label{fig-trendSPI-1}SPI (Standardized Precipitation
Index)}
\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}

\subcaption{\label{fig-trendSPI-2}SPEI (Standardized Precipitation
Evapotranspiration Index)}
\end{minipage}%
\newline
\begin{minipage}[t]{0.50\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}

\subcaption{\label{fig-trendSPI-3}EDDI (Evaporative Demand Drought
Index)}
\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{../output/figs/trend_raster_zcSM_1981-2023.png}

\subcaption{\label{fig-trendSPI-4}SSMI (Standardized Soil Moisture
Index)}
\end{minipage}%

\caption{\label{fig-trendSPI}Linear trend of the drought index (*) at
time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023}

@@

\elandscape
